---
title: "COVID metabolon QC (Xenobiotics)"
author: "Yulu Chen"
output: 
  html_document: 
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: false
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
pkg <- c("tidyverse", "data.table", "here", "readxl", "matrixStats", "e1071", "GGally", "ggrepel", "openxlsx",
         "FactoMineR", "factoextra")
for (p in pkg) {
        if (require(p, character.only = T)) {
                print(paste0(p, " loaded successfully"))
        } else {
                install.packages(p)
                require(p, character.only = T)
                print(paste0(p, " downloaded and loaded successfully"))
        }
}


## Paths

dat_dir <- "data/"
res_dir <- "results/"
fig_dir <- "figures/"

## File name

dat_fname <- str_c(dat_dir, 'NATS-01-20PHML_DATA_TABLES.xlsx')
mets_sname <- "Batch-normalized Data"
mets_info_sname <- "Chemical Annotation"
sam_info_sname <- "Sample Meta Data"

## Hard-coded numbers

cv_thld <- 0.25
missing_thld <- 0.30 
qc_rm_thld <- 0.75

## Functions

impute_halfmin <- function(x) {replace(x, is.na(x), (min(x, na.rm = T)/2))}
pareto <- function(x) {(x-mean(x, na.rm = T))/sqrt(sd(x, na.rm = T))}
```


# 1. Read data
## 1.1 Load Data
```{r}

mets <- read.xlsx(dat_fname, sheet = mets_sname) %>% data.table()
mets_info <- read.xlsx(dat_fname, sheet = mets_info_sname) %>% data.table()
sam_info <- read.xlsx(dat_fname, sheet = sam_info_sname) %>% data.table()

pheno <- sam_info

```

## 1.2 Data statistic summary
944 subjects and 1546 metabolites
```{r}
sam_info[, .N, .(GENDER)]

sam_info[, .N, .(RACE_ETHNICITY)]

summary(sam_info$AGE)

ggplot(sam_info, aes(x = AGE, color = GENDER)) + geom_histogram(fill = "white") + labs(title = "Age histogram")
```

# 2. Extract data
## 2.1 Rename Headers
```{r}

mets_info$Name <- paste0("X", mets_info$CHEM_ID)
mets_info$Names <- paste0("X", mets_info$CHEM_ID)
mets_list <- mets_info$Name 
colnames(mets) <- c("PARENT_SAMPLE_NAME", mets_list)

```

## 2.2 Combine Sample Info and Metabolites
```{r}
mets <- merge(sam_info, mets, by = "PARENT_SAMPLE_NAME") %>% data.table() # 944 1564
## c1 PARENT_SAMPLE_NAME

mets_all <- mets
```

METABOLON Batch normalized data: 
For each metabolite, the raw values in the experimental samples are divided by the median of those samples in each instrument batch, giving each batch and thus the metabolite  a median of one.  For studies containing < 144 total samples, typically no batch normalization is required and the Batch-normalized data simply reflect median-scaled raw data.  Please see "Batch and Batch Correction" section below for further details.

# 3. Xenobiotics
The super pathway information for all metabolites
```{r}
mets_info[, .N, SUPER_PATHWAY]
length(is.na(mets_info$SUPER_PATHWAY) == TRUE) # all mets has super pathway information
```
The xenobiotics information
```{r}
xenobiotics<-subset(mets_info, SUPER_PATHWAY=="Xenobiotics")
dim(xenobiotics)[1]
xenobiotics[, .N, SUB_PATHWAY]
```

# 4. Missingness
## 4.1 Missing in metabolites
all metabolites
```{r}
mets_na <- as.data.frame(matrix(ncol = 2, nrow = length(mets_list)))
colnames(mets_na) <- c("Metabolite Name", "NA percent")
mets_na$`Metabolite Name` <- mets_list

for (i in 1:length(mets_list)) {
  eval(parse(text = str_c("mets_na[i,2] <- sum(is.na(mets_all[,", mets_list[i], "]))/944")))
}

summary(mets_na$`NA percent`)
ggplot(mets_na, aes(x = `NA percent`)) + geom_histogram() + labs(title = "NA percent histogram for each metabolites")

colnames(mets_na)[1] <- "Name"
mets_info <- merge(mets_info, mets_na, by = "Name", all.x = T)
colnames(mets_na)[1] <- "Metabolite Name"
```
Xenobiotics
```{r}
xen_list <- xenobiotics$CHEM_ID
xen_list <- paste0("X", xen_list)

xen_na <- as.data.frame(matrix(ncol = 2, nrow = length(xen_list)))
colnames(xen_na) <- c("Xenobiotics Name", "NA percent")
xen_na$`Xenobiotics Name` <- xen_list

for (i in 1:length(xen_list)) {
  eval(parse(text = str_c("xen_na[i,2] <- sum(is.na(mets_all[,", xen_list[i], "]))/944")))
}

summary(xen_na$`NA percent`)
ggplot(xen_na, aes(x = `NA percent`)) + geom_histogram() + labs(title = "NA percent histogram for each Xenobiotics")

```

## 4.2 Missing in subjects
```{r}
sam_list <- sam_info$PARENT_SAMPLE_NAME

sam_na <- as.data.frame(matrix(ncol = 2, nrow = length(sam_list)))
colnames(sam_na) <- c("PARENT_SAMPLE_NAME", "NA percent")
sam_na$PARENT_SAMPLE_NAME <- sam_list

for (i in 1:length(sam_list)) {
  sam_na[i,2] <- sum(is.na(mets_all[i,c(mets_list)]))/1546
}

summary(sam_na$`NA percent`)
# ggplot(sam_na, aes(x = `NA percent`)) + geom_histogram() + labs(title = "NA percent histogram for each subject")

```

## 4.3 Remove metabolites with >= 75% missing except Xenobiotics
remove 49 metabolites and 1497 metabolites left
```{r}
## Remove metabolites with >= 75% missing (these will have post-imputation IQR = 0 anyway)
# 100% missing will cause warning when imputing with half minimum
# mets_na_75 <- mets_list[! (mets_list %in% xen_list)] #1157
# mets_na_75 <- mets_na_75[mets_na$`NA percent` >= 0.75]
mets_list_exp_xeno <- mets_list[!(mets_list %in% xen_list)]
mets_na_exp_xeno <- mets_na[!(mets_na$`Metabolite Name` %in% xen_list),]

mets_na_75 <- mets_list_exp_xeno[mets_na_exp_xeno$`NA percent`>= 0.75]
1546 - 389 - length(mets_na_75)

eval(parse(text = str_c("mets_all <- subset(mets_all, select = -c(", mets_na_75, "))")))
mets_list_qc75 <- mets_list[!(mets_list %in% mets_na_75)]
mets_list_exp_xeno_qc75 <- mets_list_exp_xeno[!(mets_list_exp_xeno %in% mets_na_75)]
```

# 5. Impute with half minimum

```{r}
mets_all[, (mets_list_qc75) := lapply(.SD, impute_halfmin), .SDcols = mets_list_qc75]
```

# 6. Pre-processing statistics & PCA
## 6.1 Pre-processing stats
```{r}

mets_pre_stats <- cbind(apply(mets_all[, ..mets_list_qc75], 2, min), 
                        apply(mets_all[, ..mets_list_qc75], 2, max), 
                        apply(mets_all[, ..mets_list_qc75], 2, IQR), 
                        apply(mets_all[, ..mets_list_qc75], 2, skewness))
mets_pre_stats <- as.data.frame(mets_pre_stats) %>% rownames_to_column() %>% as.data.table()
setnames(mets_pre_stats, c("metabolite", "min", "max", "iqr_pre", "skewness_pre"))
mets_pre_stats[, ':='(iqr_rank_pre = rank(iqr_pre, ties.method = "first"), 
                      skewness_rank_pre = rank(skewness_pre, ties.method = "first"))]

mets_pre_stats[min == max, ] # None

colnames(mets_pre_stats)[1] <- "Name"
mets_info_after_qc <- merge(mets_info, mets_pre_stats, by = "Name", all.y = T)

write.csv(mets_info_after_qc, here(res_dir, "mets_qc_infor.csv"), row.names=FALSE)

colnames(mets_pre_stats)[1] <- "metabolite"
```

IQR information
```{r}
mets_pre_stats[, .N, iqr_pre == 0]
mets_pre_stats[, summary(iqr_pre)]

ggplot(mets_pre_stats) + 
        geom_point(aes(iqr_rank_pre, iqr_pre), size = 2, alpha = 0.5) + 
        labs(title = "Pre-processing IQR (halfmin imputed)") +
        labs(x = "Rank", y = "IQR") + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))
ggsave(here(fig_dir, "iqr_preprocessing_halfmin.png"), width = 5, height = 5)

```

Skewness information
222 out of 1497 Metabolites are skewed.
```{r}
mets_pre_stats[, summary(skewness_pre)]

mets_pre_stats[, .N, skewness_pre>=2|skewness_pre<=-2]

ggplot(mets_pre_stats) + 
        geom_point(aes(skewness_rank_pre, skewness_pre), size = 2, alpha = 0.5) + 
        labs(title = "Pre-processing skewness (halfmin imputed)") +
        labs(x = "Rank", y = "Skewness") + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))
ggsave(here(fig_dir, "skewness_preprocessing_halfmin.png"), width = 5, height = 5)


```

## 6.2 Drop metabolites with IQR = 0 except Xenobiotics
217 metabolites with IQR = 0 but all these metabolites belongs to Xenobiotics so I didn't drop these metabolites
```{r}

# eval(parse(text = str_c("mets_pre_stats  <- subset(mets_pre_stats , select = -c(", xen_list, "))")))

iqr_0_list <- mets_pre_stats[mets_pre_stats$iqr_pre == 0,]
dim(iqr_0_list)[1]

iqr_0_list <- iqr_0_list$metabolite
length(intersect(iqr_0_list, xen_list))

# eval(parse(text = str_c("mets_all <- subset(mets_all, select = -c(", iqr_0_list, "))")))
# mets_list_qc75 <- mets_list_qc75[!(mets_list_qc75 %in% iqr_0_list)]

```

## 6.3 Pre-processing PCA
The PC1 explained 12.4% variance and more than 34 PC can explain more than 50% variance
```{r}

pca_pre <- prcomp(mets_all[, ..mets_list_qc75], scale. = T)
# summary(pca_pre)
pca_pre$sdev[1]^2 / sum(pca_pre$sdev^2) 
which((cumsum(pca_pre$sdev^2) / sum(pca_pre$sdev^2)) > 0.5)[1] 

fviz_eig(pca_pre)

pc_pre <- as.data.table(pca_pre$x)
pc_pre <- cbind(pc_pre, sam_info) 

ggplot(pc_pre, aes(x = PC1, y = PC2)) + 
        geom_point(size = 2, alpha = 0.5) +
        labs(title = "Pre-processing PC2 vs PC1 (halfmin imputed)") + 
        geom_label_repel(aes(label = ifelse(PC1 < -75 | PC2 > 75, PARENT_SAMPLE_NAME, "")),
                  box.padding   = 0.25, 
                  point.padding = 0.5,
                  segment.color = 'grey50') + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))
ggsave(here(fig_dir, "PC1_PC2_preprocessing_halfmin.png"))

ggplot(pc_pre, aes(x = PC1, y = PC2)) + 
        geom_point(aes(color = GENDER), size = 2, alpha = 0.5) + 
        labs(title = "Pre-processing PC2 vs PC1 (halfmin imputed)") + 
        geom_label_repel(aes(label = ifelse(PC1 < -75 | PC2 > 75, PARENT_SAMPLE_NAME, "")),
                  box.padding   = 0.25, 
                  point.padding = 0.5,
                  segment.color = 'grey50') + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))

ggsave(here(fig_dir, "PC1_PC2_preprocessing_by_gender_halfmin.png"))

ggplot(pc_pre, aes(x = PC1, y = PC2)) + 
        geom_point(aes(color = AGE), size = 2, alpha = 0.5) + 
        labs(title = "Pre-processing PC2 vs PC1 (halfmin imputed)") + 
        geom_label_repel(aes(label = ifelse(PC1 < -75 | PC2 > 75, PARENT_SAMPLE_NAME, "")),
                  box.padding   = 0.25, 
                  point.padding = 0.5,
                  segment.color = 'grey50') + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))

ggsave(here(fig_dir, "PC1_PC2_preprocessing_by_age_halfmin.png"))

ggplot(pc_pre, aes(x = PC1, y = PC2)) + 
        geom_point(aes(color = RACE_ETHNICITY), size = 2, alpha = 0.5) + 
        labs(title = "Pre-processing PC2 vs PC1 (halfmin imputed)") + 
        geom_label_repel(aes(label = ifelse(PC1 < -75 | PC2 > 75, PARENT_SAMPLE_NAME, "")),
                  box.padding   = 0.25, 
                  point.padding = 0.5,
                  segment.color = 'grey50') + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))

ggsave(here(fig_dir, "PC1_PC2_preprocessing_by_race_halfmin.png"))
```

# 7. Log-transformation & pareto-scaling
```{r}

## log 10 transform all metabolites

mets_all[, (mets_list_qc75) := log10(.SD + 1), .SDcols = mets_list_qc75]

## Pareto-scale the features
## Pareto-scaling = mean-centered and divided by the square root of standard deviation 

# mets_all[, min(.SD), .SDcols = mets_list_qc75] #2.743995e-05
# mets_all[, max(.SD), .SDcols = mets_list_qc75] #4.736254

mets_all[, (mets_list_qc75) := lapply(.SD, pareto), .SDcols = mets_list_qc75]

```

# 8. Post-processing statistics & PCA
## 8.1 Post-processing PCA (include xeno)
The PC1 explained 12.9% variance and more than 25 PC can explain more than 50% variance
```{r}
mets_final_df <- as.data.frame(mets_all)
rownames(mets_final_df) <- mets_final_df$PARENT_SAMPLE_NAME
mets_final_df <- subset(mets_final_df, select = mets_list_qc75)

pca_post <- prcomp(mets_final_df, scale. = T)

fviz_eig(pca_post)

pca_post$sdev[1]^2 / sum(pca_post$sdev^2) 
which((cumsum(pca_post$sdev^2) / sum(pca_post$sdev^2)) > 0.5)[1] 

pc_post <- pca_post$x %>% as.data.frame()

ggplot(pc_post, aes(x = PC1, y = PC2)) + 
        geom_point(size = 2, alpha = 0.5) + 
        labs(title = "Post-processing PC2 vs PC1 (halfmin imputed)") + 
        geom_label_repel(aes(label = ifelse(PC1 < -75 | PC2 > 75, rownames(pc_post), "")),
                  box.padding   = 0.25, 
                  point.padding = 0.5,
                  segment.color = 'grey50') + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))
ggsave(here(fig_dir, "PC1_PC2_postprocessing_halfmin.png"))

ggpairs(pc_post[, 1:10], aes(alpha = 0.1)) + 
        labs(title = "Post-processing PC1-10 pairwise plots (halfmin imputed)") + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))
ggsave(here(fig_dir, "PC1_PC10_pairs_postprocessing_halfmin.png"), width = 15, height = 12)


```
## 8.2. Post-processing stats (include xeno)
IQR information
```{r}
mets_post_stats <- cbind(apply(mets_all[, ..mets_list_qc75], 2, IQR), 
                         apply(mets_all[, ..mets_list_qc75], 2, skewness))
mets_post_stats <- as.data.frame(mets_post_stats) %>% rownames_to_column() %>% as.data.table()
setnames(mets_post_stats, c("metabolite", "iqr_post", "skewness_post"))
mets_post_stats[, ':='(iqr_rank_post = rank(iqr_post, ties.method = "first"), 
                       skewness_rank_post = rank(skewness_post, ties.method = "first"))]

mets_post_stats[, summary(iqr_post)]

ggplot(mets_post_stats) + 
        geom_point(aes(iqr_rank_post, iqr_post), size = 2, alpha = 0.5) + 
        labs(title = "Post-processing IQR (halfmin imputed)") +
        labs(x = "Rank", y = "IQR") + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))
ggsave(here(fig_dir, "iqr_postprocessing_halfmin.png"), width = 5, height = 5)
```
Skewness information
498 out of 1497 Metabolites are skewed.
```{r}
mets_post_stats[, summary(skewness_post)]

mets_post_stats[, .N, skewness_post>=2|skewness_post<=-2]


ggplot(mets_post_stats) + 
        geom_point(aes(skewness_rank_post, skewness_post), size = 2, alpha = 0.5) + 
        labs(title = "Post-processing skewness (halfmin imputed)") +
        labs(x = "Rank", y = "Skewness") + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))
ggsave(here(fig_dir, "skewness_postprocessing_halfmin.png"), width = 5, height = 5)

```


## 8.3 Post-processing PCA (exclude xeno)
The PC1 explained 15.9% variance and more than 15 PC can explain more than 50% variance
```{r}
mets_final_df_exp_xeno <- as.data.frame(mets_all)
rownames(mets_final_df_exp_xeno) <- mets_final_df_exp_xeno$PARENT_SAMPLE_NAME
mets_final_df_exp_xeno <- subset(mets_final_df_exp_xeno, select = mets_list_exp_xeno_qc75)

pca_post <- prcomp(mets_final_df_exp_xeno, scale. = T)

fviz_eig(pca_post)

pca_post$sdev[1]^2 / sum(pca_post$sdev^2) 
which((cumsum(pca_post$sdev^2) / sum(pca_post$sdev^2)) > 0.5)[1] 

pc_post <- pca_post$x %>% as.data.frame()

ggplot(pc_post, aes(x = PC1, y = PC2)) + 
        geom_point(size = 2, alpha = 0.5) + 
        labs(title = "Post-processing PC2 vs PC1 (halfmin imputed)") + 
        geom_label_repel(aes(label = ifelse(PC1 < -75 | PC2 > 75, rownames(pc_post), "")),
                  box.padding   = 0.25, 
                  point.padding = 0.5,
                  segment.color = 'grey50') + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))
ggsave(here(fig_dir, "PC1_PC2_postprocessing_halfmin_exp_xeno.png"))

ggpairs(pc_post[, 1:10], aes(alpha = 0.1)) + 
        labs(title = "Post-processing PC1-10 pairwise plots (halfmin imputed)") + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))
ggsave(here(fig_dir, "PC1_PC10_pairs_postprocessing_halfmin_exp_xeno.png"), width = 15, height = 12)


```
## 8.4 Post-processing stats (exclude xeno)
IQR information
```{r}
mets_post_stats <- cbind(apply(mets_all[, ..mets_list_exp_xeno_qc75], 2, IQR), 
                         apply(mets_all[, ..mets_list_exp_xeno_qc75], 2, skewness))
mets_post_stats <- as.data.frame(mets_post_stats) %>% rownames_to_column() %>% as.data.table()
setnames(mets_post_stats, c("metabolite", "iqr_post", "skewness_post"))
mets_post_stats[, ':='(iqr_rank_post = rank(iqr_post, ties.method = "first"), 
                       skewness_rank_post = rank(skewness_post, ties.method = "first"))]

mets_post_stats[, summary(iqr_post)]

ggplot(mets_post_stats) + 
        geom_point(aes(iqr_rank_post, iqr_post), size = 2, alpha = 0.5) + 
        labs(title = "Post-processing IQR (halfmin imputed)") +
        labs(x = "Rank", y = "IQR") + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))
ggsave(here(fig_dir, "iqr_postprocessing_halfmin_exp_xeno.png"), width = 5, height = 5)
```
Skewness information
```{r}
mets_post_stats[, summary(skewness_post)]

mets_post_stats[, .N, skewness_post>=2|skewness_post<=-2]


ggplot(mets_post_stats) + 
        geom_point(aes(skewness_rank_post, skewness_post), size = 2, alpha = 0.5) + 
        labs(title = "Post-processing skewness (halfmin imputed)") +
        labs(x = "Rank", y = "Skewness") + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))
ggsave(here(fig_dir, "skewness_postprocessing_halfmin_exp_xeno.png"), width = 5, height = 5)

```

# 9. PC plots by sample characteristics
PCA expect for Xenobiotics

The PC1 explained 15.9% variance and more than 15 PCs can explain more than 50% variance
```{r}
# pca except xenobiotics
mets_final_df_exp_xeno <- subset(mets_final_df, select = mets_list_exp_xeno_qc75)
# mets_final_df_exp_xeno[1:10,1:10]

pca_post_exp_xeno <- prcomp(mets_final_df_exp_xeno, scale. = T)

fviz_eig(pca_post_exp_xeno)

pca_post_exp_xeno$sdev[1]^2 / sum(pca_post_exp_xeno$sdev^2) 
which((cumsum(pca_post_exp_xeno$sdev^2) / sum(pca_post_exp_xeno$sdev^2)) > 0.5)[1] 

pc_post_exp_xeno <- pca_post_exp_xeno$x %>% as.data.frame()

```

```{r}
pc_post <- cbind(pc_post, sam_info) 
pc_post_exp_xeno <- cbind(pc_post_exp_xeno, sam_info)

ggplot(pc_post, aes(x = PC1, y = PC2)) + 
        geom_point(aes(color = GENDER), size = 2, alpha = 0.5) + 
        labs(title = "Post-processing PC2 vs PC1 (halfmin imputed)") + 
        geom_label_repel(aes(label = ifelse(PC1 < -75 | PC2 > 75, PARENT_SAMPLE_NAME, "")),
                  box.padding   = 0.25, 
                  point.padding = 0.5,
                  segment.color = 'grey50') + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))

ggsave(here(fig_dir, "PC1_PC2_postprocessing_by_gender_halfmin.png"))

ggplot(pc_post_exp_xeno, aes(x = PC1, y = PC2)) + 
        geom_point(aes(color = GENDER), size = 2, alpha = 0.5) + 
        labs(title = "Post-processing PC2 vs PC1 (halfmin imputed, except for Xenobiotics)") + 
        geom_label_repel(aes(label = ifelse(PC1 < -75 | PC2 > 75, PARENT_SAMPLE_NAME, "")),
                  box.padding   = 0.25, 
                  point.padding = 0.5,
                  segment.color = 'grey50') + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))

ggsave(here(fig_dir, "PC1_PC2_postprocessing_by_gender_halfmin_exp_xeno.png"))


ggplot(pc_post, aes(x = PC1, y = PC2)) + 
        geom_point(aes(color = AGE), size = 2, alpha = 0.5) + 
        labs(title = "Post-processing PC2 vs PC1 (halfmin imputed)") + 
        geom_label_repel(aes(label = ifelse(PC1 < -75 | PC2 > 75, PARENT_SAMPLE_NAME, "")),
                  box.padding   = 0.25, 
                  point.padding = 0.5,
                  segment.color = 'grey50') + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))

ggsave(here(fig_dir, "PC1_PC2_postprocessing_by_age_halfmin.png"))

ggplot(pc_post_exp_xeno, aes(x = PC1, y = PC2)) + 
        geom_point(aes(color = AGE), size = 2, alpha = 0.5) + 
        labs(title = "Post-processing PC2 vs PC1 (halfmin imputed, except for Xenobiotics)") + 
        geom_label_repel(aes(label = ifelse(PC1 < -75 | PC2 > 75, PARENT_SAMPLE_NAME, "")),
                  box.padding   = 0.25, 
                  point.padding = 0.5,
                  segment.color = 'grey50') + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))

ggsave(here(fig_dir, "PC1_PC2_postprocessing_by_age_halfmin_exp_xeno.png"))


ggplot(pc_post, aes(x = PC1, y = PC2)) + 
        geom_point(aes(color = RACE_ETHNICITY), size = 2, alpha = 0.5) + 
        labs(title = "Post-processing PC2 vs PC1 (halfmin imputed)") + 
        geom_label_repel(aes(label = ifelse(PC1 < -75 | PC2 > 75, PARENT_SAMPLE_NAME, "")),
                  box.padding   = 0.25, 
                  point.padding = 0.5,
                  segment.color = 'grey50') + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))

ggsave(here(fig_dir, "PC1_PC2_postprocessing_by_race_halfmin_exp_xeno.png"))

ggplot(pc_post_exp_xeno, aes(x = PC1, y = PC2)) + 
        geom_point(aes(color = RACE_ETHNICITY), size = 2, alpha = 0.5) + 
        labs(title = "Post-processing PC2 vs PC1 (halfmin imputed, except for Xenobiotics)") + 
        geom_label_repel(aes(label = ifelse(PC1 < -75 | PC2 > 75, PARENT_SAMPLE_NAME, "")),
                  box.padding   = 0.25, 
                  point.padding = 0.5,
                  segment.color = 'grey50') + 
        theme_minimal() + 
        theme(plot.title = element_text(size = 12, face = "bold"), 
              axis.text = element_text(size = 12), 
              axis.title = element_text(size = 12))

```

# 10. Save data
```{r}
write.csv(mets_final_df, here(res_dir, "Mets_Global_QC_version2.csv"), row.names=T)

save(mets_final_df, mets_final_df_exp_xeno, mets_list_qc75, mets_list_exp_xeno_qc75, mets_list, mets_na_75, mets_list_exp_xeno, xen_list, mets_info, mets_na, sam_info, 
     file = here(res_dir, "processed_QC_version2.RData"))
```

# Session info
```{r}
sessionInfo()
```


