---
title: "COVID: SNF parameter selection"
subtitle: "subjects with plasma before covid, mets except xenobiotics"
author: "Yulu Chen"
output: 
  html_document: 
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: false
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

# 1. Setup
## 1.1 Load packages
```{r, include=FALSE}
## Packages

pkg <- c("data.table", "plyr", "igraph", "SNFtool", "ggpubr", "RColorBrewer", "lattice", "dplyr", "stringr", "qgraph", "ggalluvial", "tidyverse", "DT", "GGally", "here", "table1", "ggrepel", "openxlsx",
         "MASS", "Hmisc")
for (p in pkg) {
        if (require(p, character.only = T)) {
                print(paste0(p, " loaded successfully"))
        } else {
                install.packages(p)
                require(p, character.only = T)
                print(paste0(p, " downloaded and loaded successfully"))
        }
}

```

## 1.2 Functions
```{r}
# Diagonals for heatmap set to 0
diag0 <- function(df){
    df2 <-df
    diag(df2) <- 0
    return(df2)
}

```

## 1.3 Set Paths
```{r}
dat_dir <- "data/"
fig_dir <- "figures"
res_dir <- "results"
```

## 1.4 Import Data
Import metabolone data and phenotype data
```{r}
# load QC metabolomics data
load("~/results/processed_QC_version2.RData",  verbose = T)

# load meta data
pheno <- read.xlsx("~/data/LEOCC_meta_data.xlsx", 1)

pheno_disease <- fread("~/data/LEOCC_pheno_disease.csv")

pheno_immune <- read.table(file = here(dat_dir, "pheno_immune.tsv"), sep = '\t', header = TRUE)

pheno_flu <- fread("~/data/pheno_flu.csv")

pheno_lab_test <- fread("~/data/pheno_lab_test.csv")

```


## 1.5 Select the Subject with Pre-covid Plasma Samples
444 subjects with pre-covid plasma sample only
475 subjects with pre-covid plasma sample
```{r}
subject_plasma_before_covid <- pheno[sample_term == "before COVID",]

```

## 1.6 Match Samples in Mets, Sam and Pheno Info
```{r}

# merge sam info and pheno
subject_plasma_before_covid_id <- subject_plasma_before_covid$BiobankSubjectID

any(duplicated(subject_plasma_before_covid_id)) #FALSE

sam_df <- sam_info[sam_info$BiobankSubjectID %in% subject_plasma_before_covid_id,] #520
any(duplicated(sam_df$BiobankSubjectID)) #TRUE
any(is.na(sam_df$Confirmed_Date)) #FALSE

sam_df <- merge(before_id, sam_df, by = "CLIENT_IDENTIFIER")
any(is.na(sam_df$BiobankSubjectID.x))
any(is.na(sam_df$BiobankSubjectID.y))
sum(is.na(sam_df$BiobankSubjectID.y))

sam_df_id <- sam_df[,c(1,2,4,20)]
colnames(sam_df_id)[2] <- "BiobankSubjectID"

mets <- mets_final_df %>% rownames_to_column()
colnames(mets)[1] <- c("PARENT_SAMPLE_NAME")
mets[1:10,1:10]

mets <- merge(sam_df_id, mets, by = "PARENT_SAMPLE_NAME")
# mets <- mets[mets$PARENT_SAMPLE_NAME %in% sam_df_id$PARENT_SAMPLE_NAME,]
mets[1:10,1:10]
mets <- mets[,-c(1,2,4)]

dim(mets) # 474 1498
mets[1:10,1:10]

mets$BiobankSubjectID <- as.integer(mets$BiobankSubjectID)

## Merge pheno and mets data
df <- merge(subject_plasma_before_covid, mets, by = "BiobankSubjectID")
dim(df)
df[1:10,1:20]

pheno_disease <- pheno_disease[,-c(2:6)]
colnames(pheno_disease)[1] <- "BiobankSubjectID"
df <- merge(df, pheno_disease, by = "BiobankSubjectID", all.x = T)

pheno_flu <- pheno_flu[,-c(2:6)]
colnames(pheno_flu)[1] <- "BiobankSubjectID"
df <- merge(df, pheno_flu, by = "BiobankSubjectID", all.x = T)

colnames(lab_test_df)[1] <- "BiobankSubjectID"
df <- merge(df, lab_test_df, by = "BiobankSubjectID", all.x = T)
```

## 1.7 Basic statistic summary
```{r}
# race
df$Race[df$Race == "Asian"] <- "Other"
df$Race[df$Race == "Unknown"] <- "Other"
df[, .N, Race][, pct := N/sum(N)*100] %>% print(digits = 3)

# ethnicity
df[, .N, Ethnicity][, pct := N/sum(N)*100] %>% print(digits = 3)

# gender
df[, .N, Gender][, pct := N/sum(N)*100] %>% print(digits = 3)

# smoking status
df$hasSmokingBeforeCovid <- as.character(as.integer(df$hasSmokingBeforeCovid))
class(df$hasSmokingBeforeCovid)
df[, .N, hasSmokingBeforeCovid][, pct := N/sum(N)*100] %>% print(digits = 3)

# BMI
ggplot(df, aes(x = BMI_med)) + geom_histogram() + labs(title = "BMI Median histogram")
df[, summary(BMI_med)]
df[, sd(BMI_med)]

# plasma sample collection age
df[, summary(Age)]
df[, sd(Age)]

# deceased
df[, .N, Deceased][, pct := N/sum(N)*100] %>% print(digits = 3)

```

## 1.8 Estimate residual of metabolites
455 subjects were useed to further analysis

model: Mets ~ Gender + Race + Ethnicity + Smoking + Age + BMI

all platforms
```{r}
# QC mets for all mets
df_res <- data.frame(cbind(df$BiobankSubjectID, df$BMI_med))
colnames(df_res) <- c("BiobankSubjectID", "BMI_med")

for (met in mets_list_qc75) {
  fit <- eval(parse(text = str_c("glm(", met, " ~ Gender + Race + Ethnicity + hasSmokingBeforeCovid + Age + BMI_med, data = df)")))
  df_res[,met] <- resid(fit, na.action=na.exclude)
}

dim(df_res)
df_res[1:10,1:10]

# QC mets except for xeno
df_res_exp_xeno <- df_res[,c("BiobankSubjectID","BMI_med", mets_list_exp_xeno_qc75)]
dim(df_res_exp_xeno)
df_res_exp_xeno[1:10,1:10]

```

mets list
```{r}
# metabolomics list (missing percent less than 75%)
mets_info_qc75 <- mets_info[mets_info$Name %in% mets_list_qc75, ]
table(mets_info_qc75$PLATFORM)

# metabolomics list except for xenobiotics (missing percent less than 75%)
mets_info_qc75_exp_xeno <- mets_info[mets_info$Name %in% mets_list_exp_xeno_qc75, ]
table(mets_info_qc75_exp_xeno$PLATFORM)
```

Neg platform
```{r}
# QC mets
df_res_neg <- data.frame(cbind(df$BiobankSubjectID, df$BMI_med))
colnames(df_res_neg) <- c("BiobankSubjectID", "BMI_med")
df_res_neg <- df_res_neg[complete.cases(df_res_neg),]

mets_list_qc75_neg <- mets_info_qc75$Name[mets_info_qc75$PLATFORM == "Neg"]
for (met in mets_list_qc75_neg) {
  fit <- eval(parse(text = str_c("glm(", met, " ~ Gender + Race + Ethnicity + hasSmokingBeforeCovid + Age + BMI_med, data = df)")))
  df_res_neg[,met] <- resid(fit, na.action=na.exclude)
}

dim(df_res_neg)
df_res_neg[1:10,1:10]

# QC mets except for xeno
mets_list_qc75_exp_xeno_neg <- mets_info_qc75_exp_xeno$Name[mets_info_qc75_exp_xeno$PLATFORM == "Neg"]
df_res_exp_xeno_neg <- df_res_neg[,c("BiobankSubjectID","BMI_med", mets_list_qc75_exp_xeno_neg)]

```

Polar platform
```{r}
# QC mets
df_res_polar <- data.frame(cbind(df$BiobankSubjectID, df$BMI_med))
colnames(df_res_polar) <- c("BiobankSubjectID", "BMI_med")
df_res_polar <- df_res_polar[complete.cases(df_res_polar),]

mets_list_qc75_polar <- mets_info_qc75$Name[mets_info_qc75$PLATFORM == "Polar"]
for (met in mets_list_qc75_polar) {
  fit <- eval(parse(text = str_c("glm(", met, " ~ Gender + Race + Ethnicity + hasSmokingBeforeCovid + Age + BMI_med, data = df)")))
  df_res_polar[,met] <- resid(fit, na.action=na.exclude)
}

dim(df_res_polar)
df_res_polar[1:10,1:10]

# QC mets except for xeno
mets_list_qc75_exp_xeno_polar <- mets_info_qc75_exp_xeno$Name[mets_info_qc75_exp_xeno$PLATFORM == "Polar"]
df_res_exp_xeno_polar <- df_res_polar[,c("BiobankSubjectID","BMI_med", mets_list_qc75_exp_xeno_polar)]

```

Pos Early platform
```{r}
# QC mets
df_res_posearly <- data.frame(cbind(df$BiobankSubjectID, df$BMI_med))
colnames(df_res_posearly) <- c("BiobankSubjectID", "BMI_med")
df_res_posearly <- df_res_posearly[complete.cases(df_res_posearly),]

mets_list_qc75_posearly <- mets_info_qc75$Name[mets_info_qc75$PLATFORM == "Pos Early"]
for (met in mets_list_qc75_posearly) {
  fit <- eval(parse(text = str_c("glm(", met, " ~ Gender + Race + Ethnicity + hasSmokingBeforeCovid + Age + BMI_med, data = df)")))
  df_res_posearly[,met] <- resid(fit, na.action=na.exclude)
}

dim(df_res_posearly)
df_res_posearly[1:10,1:10]

# QC mets except for xeno
mets_list_qc75_exp_xeno_posearly <- mets_info_qc75_exp_xeno$Name[mets_info_qc75_exp_xeno$PLATFORM == "Pos Early"]
df_res_exp_xeno_posearly <- df_res_posearly[,c("BiobankSubjectID","BMI_med", mets_list_qc75_exp_xeno_posearly)]

```


Pos Late platform
```{r}
# QC mets
df_res_poslate <- data.frame(cbind(df$BiobankSubjectID, df$BMI_med))
colnames(df_res_poslate) <- c("BiobankSubjectID", "BMI_med")
df_res_poslate <- df_res_poslate[complete.cases(df_res_poslate),]

mets_list_qc75_poslate <- mets_info_qc75$Name[mets_info_qc75$PLATFORM == "Pos Late"]
for (met in mets_list_qc75_poslate) {
  fit <- eval(parse(text = str_c("glm(", met, " ~ Gender + Race + Ethnicity + hasSmokingBeforeCovid + Age + BMI_med, data = df)")))
  df_res_poslate[,met] <- resid(fit, na.action=na.exclude)
}

dim(df_res_poslate)
df_res_poslate[1:10,1:10]

# QC mets except for xeno
mets_list_qc75_exp_xeno_poslate <- mets_info_qc75_exp_xeno$Name[mets_info_qc75_exp_xeno$PLATFORM == "Pos Late"]
df_res_exp_xeno_poslate <- df_res_poslate[,c("BiobankSubjectID","BMI_med", mets_list_qc75_exp_xeno_poslate)]


```



# 2. Similarity Network Fusion
## 2.1 Calculate Distance Matrices
computes the squared Euclidean distances between all pairs of data points
```{r}
# QC mets
df_res_exp_xeno<- df_res_exp_xeno[,-c(1,2)]
Dist <- (dist2(as.matrix(df_res_exp_xeno),as.matrix(df_res_exp_xeno)))^(1/2)
dim(Dist)

df_res_exp_xeno_neg<- df_res_exp_xeno_neg[,-c(1,2)]
df_res_exp_xeno_polar<- df_res_exp_xeno_polar[,-c(1,2)]
df_res_exp_xeno_posearly<- df_res_exp_xeno_posearly[,-c(1,2)]
df_res_exp_xeno_poslate<- df_res_exp_xeno_poslate[,-c(1,2)]

Dist_neg = (dist2(as.matrix(df_res_exp_xeno_neg),as.matrix(df_res_exp_xeno_neg)))^(1/2)
Dist_polar = (dist2(as.matrix(df_res_exp_xeno_polar),as.matrix(df_res_exp_xeno_polar)))^(1/2)
Dist_posearly = (dist2(as.matrix(df_res_exp_xeno_posearly),as.matrix(df_res_exp_xeno_posearly)))^(1/2)
Dist_poslate = (dist2(as.matrix(df_res_exp_xeno_poslate),as.matrix(df_res_exp_xeno_poslate)))^(1/2)

```


## 2.2 Test tuning parameters, output 3 best combinations (optimize variance)
```{r}
## Only run this once -- takes a long time to run
filtmetsEuclidDistList <- list(Dist_neg, Dist_polar, Dist_posearly, Dist_poslate)
for(i in 1:4){
    varResultsDf <- data.frame(KNN=integer(),alpha=numeric(),variance=numeric())
    for(K in seq(20,150,1)){
        for(alpha in seq(0.3,1,0.1)){
            ## Draw similarity graphs
            simGraph <- affinityMatrix(filtmetsEuclidDistList[[i]], K, alpha)
            ## Calculate the variance of the upper triangle
            upperTriVar <- var(simGraph[upper.tri(simGraph)])
            ## Build results data frame
            varResultsDf <- rbind(varResultsDf,
                                  data.frame(KNN=K,alpha=alpha,
                                             variance=upperTriVar))
        }
    }
    ## Report results
    topRes <- varResultsDf[order(-varResultsDf$variance),][1:3,]
    print(topRes)
    ## Show graphs for the top 3 highest variance combinations
    for(j in 1:nrow(topRes)){
      simGraph <- affinityMatrix(filtmetsEuclidDistList[[i]], K=topRes$KNN[j], sigma=topRes$alpha[j])
      diag(simGraph) <- 0 # for plotting diag well
      distNorm <- function(x){
        x <- x/rowSums(x)
        x <- x + t(x)
        dist(x)
      }
      heatmap(x=simGraph,col=rev(colorRampPalette(brewer.pal(10, "RdBu"))(256)),
              scale="none",distfun=distNorm,symm=TRUE,
              hclust=function(x) hclust(x,method="complete"))
    }
}


```


## 2.3 Test tuning parameters when K is stable, output 3 best combinations (optimize variance)
```{r}
## Only run this once -- takes a long time to run
filtmetsEuclidDistList <- list(Dist_neg, Dist_polar, Dist_posearly, Dist_poslate)
for(i in 1:4){
    varResultsDf <- data.frame(KNN=integer(),alpha=numeric(),variance=numeric())
        for(alpha in seq(0.3,1,0.1)){
            ## Draw similarity graphs
            simGraph <- affinityMatrix(filtmetsEuclidDistList[[i]], 46, alpha)
            ## Calculate the variance of the upper triangle
            upperTriVar <- var(simGraph[upper.tri(simGraph)])
            ## Build results data frame
            varResultsDf <- rbind(varResultsDf,
                                  data.frame(KNN=46,alpha=alpha,
                                             variance=upperTriVar))
        }

    ## Report results
    topRes <- varResultsDf[order(-varResultsDf$variance),][1:3,]
    print(topRes)
    ## Show graphs for the top 3 highest variance combinations
    for(j in 1:nrow(topRes)){
      simGraph <- affinityMatrix(filtmetsEuclidDistList[[i]], K=topRes$KNN[j], sigma=topRes$alpha[j])
      diag(simGraph) <- 0 # for plotting diag well
      distNorm <- function(x){
        x <- x/rowSums(x)
        x <- x + t(x)
        dist(x)
      }
      heatmap(x=simGraph,col=rev(colorRampPalette(brewer.pal(10, "RdBu"))(256)),
              scale="none",distfun=distNorm,symm=TRUE,
              hclust=function(x) hclust(x,method="complete"))
    }
}


```

# 3. Session info
```{r}
sessionInfo()
```
